---
title: "R Intro Exercises"
output: 
  learnr::tutorial:
    theme: readable
    ace_theme: github
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome!

This tutorial will give you the chance to practice running R code without 
having to worry about the technical details of getting R running on your own
computer. Of course, it will be important to start using R on your computer,
but I think you'll feel more confident navigating any obstacles to setup after
you've successfully practiced writing and running some actual R code. Here goes:

## Functions: Something In, Something Out 

The only thing R does is turn input into output. It does this in many creative,
complex ways, but deep down it's as simple as **"Something in, something out."**

### Print

To see what I mean, run the code below. Either click the "**Run Code**" button
or click somewhere within the text of the code and press `Ctrl` + `Enter`. 

```{r print, exercise=TRUE}
print("Hello, World!")
```

You just ran a function using R code! You provided the **input** "Hello, World!" to the `print()` function and saw
the **output** printed to the screen (not to a printer). In this case, the 
output was identical to the input. You might wonder why a function exists that 
spits out exactly what it takes in. Hold that thought and run the code below:

```{r addition, exercise=TRUE}
print(60 * (24 - 7.5))
```

```{r quiz-1}
quiz(
  question("Think carefully about why that input turned into that output.
           *Specifically, if `print()` just prints your input to the screen, 
           why wasn't the output* \"60 * (24 - 7.5))\"? Choose the most accurate
           answer.",
    answer("A) Because R anticipates what you want, so when it sees math, 
           it calculates and prints the answer.",
           message = "This was tricky. R does end up solving the math problem, 
           but not because R knows what you want. R has absolutely no
           common sense and will only do what you explicitly tell it to do."),
    answer("B) Because we didn't put quotes (\") around our input.",
           message = "This was tricky. 
           It's true that if we put quotes around our input it 
           would have printed verbatim, but that's not the whole picture
           (we'll discuss more later)."),
    answer("C) Because `+` and `*` are functions, so they transform their input
           and provide their output before `print()` performs its own work",
           correct = TRUE),
    allow_retry = TRUE,
    try_again = "Try this until you get it right.",
    correct = "Right! Tricky because they look different from most functions
    (read more about this next)."
  )
)
```

### Seriously, *Everything* is a Function

The last question was kind of unfair because most functions look like
`function_name(input)` (e.g., `print("something")`), but those math functions
don't. But remember I told you that everything in R happens because functions
accept input and provide output. That's true even when something doesn't
have that classic function look.

To make this clearer, let's re-write those mathematical functions as if they
were normal functions:

```{r math-fun, exercise=TRUE}
# When we're using a special symbol as the name of a function, 
# we have to enclose it in backticks (``).
`-`(24, 7.5)
```

Subtraction is just a function with two inputs—also called 
arguments. The function subtracts the second argument
from the first.

Similarly, for multiplication:

```{r math-fun2, exercise=TRUE}
`*`(60, 16.5)
```

<!-- To take this to an absurd extreme, the same thing is true for the parentheses  -->
<!-- we're using to define the order of operations: -->

<!-- ```{r paren-fun, exercise=TRUE} -->
<!-- `*`( 60, `(`( `-`( 24, 7.5  )) ) -->
<!-- ``` -->

<!-- Okay, we got the same answer but that's super confusing. 
-->

Why am I telling you this? Because I want to drill it into your head that
everything that happens in R is the result of functions taking inputs and
creating outputs. 

### Directing Functions' Output

But how did those answers print to the screen even though we didn't use the
`print()` function? Okay, I know I just told you that R only does what you
explicitly tell it to and this seems like R is making an assumption about 
what we want to happen when we run some code. Well, R always does *something*
with the result of any code you run, and the default action is to print it.

Unless you tell it to do something else with the result, R prints it to the
screen once it calculates a result. But even then, it's the `print()` function
making it happen. Behind the scenes, `print()` takes in the result of your code
and outputs it to the screen as text. **Something in, something out**.

So, what else could we do with the output from a function? We've already seen
one subtle example: when we ran this: `print(60 * (24 - 7.5))`, we were really
putting one function (subtraction) inside another function (multiplication)
inside another function (`print()`). The **output** of the subtraction function
gets used as one **input** to the multiplication function, and the **output**
from that function gets used as the **input** to the `print()` function. 

### Nesting Functions

This idea of one function within another function is really common, but you can
see how it gets confusing really quickly. When you have functions inside 
functions inside functions, you basically have to read code from the inside out
to understand what it's doing. Later we'll discuss dealing with this, but
for now understand that putting **function A** inside **function B** is one way
of directing the output of function A to be used as an input for function B.
Read on!

## Packages Are Collections of Functions

Enough silly theory. Now that you are sick of hearing about functions, let's 
put some functions to good use.

One of the most common functions you will use in R is the `library()` function.
The name of an R package goes between the parentheses, with or without quotes.
What does this do? Well, it's a bit like wishing for more wishes: it is used 
to load packages, which are really just collections
(libraries, if you will) of functions that aren't available until you make 
them available. 

Base R provides a number of functions that can do some 
pretty useful things. These functions are made available automatically when
you start R. In addition, R users and developers have written functions
that either improve the usability of base R functions or provide capabilities
that aren't built into R (e.g., specialized statistical techniques or plotting
tools).

### The `tidyverse`

The tidyverse is a collection of several packages 
written for specific purposes but with common design principles. 
This means that as you learn the functions from one tidyverse package, 
you'll start to intuitively recognize how to use functions from the other
tidyverse packages.

Try loading the tidyverse by modifying the code below (click the "Hint" button
if you get stuck):

```{r library-fun, exercise=TRUE}
library()
```

```{r library-fun-hint}
library(tidyverse)
```

The argument (input) we provided to the `library()` function was the name of
a package. The output is a message letting us know that we actually loaded 
several packages at once that are all collected under that one name. It also
informs us that the names of a few functions from those packages conflict with
names of functions from base R. That shouldn't be a problem for now, but be 
aware that it can happen.

### Browsing Available Functions

After you load a package with the `library()` function, all functions within 
that package will be available for you to use. One way to see some of those 
functions is to type the name of a package, followed by two colons (`::`). 
If you are in RStudio (or this tutorial), autocomplete should show you a list 
of functions available in that package. 

Try it! Since we loaded the `stringr` package with the rest of the tidyverse,
try browsing the list of functions in that package below. **Hint: some packages
contain additional resources, such as datasets. You should see both when browsing
a package's contents. Try `stringr::fruit` to see a collection of character
strings made available for learning the `stringr` package's functions.

```{r browse-fun, exercise=TRUE}
stringr
```

### Example: the Lubridate Package

Let's get a quick introduction to using functions. For practice, we'll use
some date and time related functions from the `lubridate` package. 

First, complete the code below to load the `lubridate` package (remember which function to use? If not, click "Hint"):

```{r load-lubridate, exercise=TRUE}
(lubridate)
```

```{r load-lubridate-hint}
library(lubridate)
```

This won't be a full worked example, but to make it concrete let's think about
some of the things we would need to do to calculate a patient's age so we could
use that value in a report. 

One thing we need to do is get R to recognize the date string in our data
as a Date object. This is important so R knows how to do math with it.

Say the patient's date of birth is March 15, 1925, and it is stored in our 
dataset as "4/15/1925". We can use the `mdy()` function from `lubridate` to 
convert that input to a date output:

```{r setupLub}
library(lubridate)
```


```{r mdy-ex, exercise=TRUE, exercise.setup="setupLub"}
mdy("4/15/1925")
```

The output from that doesn't look fancy, but it tells us that R is able to 
transform that string into a machine-readable date object.

Next, we need to capture today's date (assuming we're calculating a patient's 
age as of today). We could just do the same thing as above but substitute 
today's date. But if we're interested in turning this into something that we can
re-use without remembering to manually update values, we need a **function**
that can give us today's date. Here it is:

```{r sysdate, exercise=TRUE, exercise.setup="setupLub"}
today()
```

Knowing what you know, can you use those functions to calculate that patient's 
age?

```{r calc-age, exercise=TRUE, exercise.setup="setupLub"}

```

```{r calc-age-hint, eval=FALSE}
today() - mdy("4/5/1925")
```

There are a few more steps to calculate it in years, etc. but we'll stop there.

## Objects/Variables

Since we're now thinking in terms of "function A, function B", let's go a step
further and start saving the output of one function so we can use it more 
easily in another function.

Here's how I would capture the output of a mathematical function to use as input
for another function later:

```{r, echo=TRUE}
waking_hours <- 24 - 7.5
waking_hours
```

The first line of code above executes the subtraction function and then 
associates the result with an R object called `waking_hours`. The `<-` symbol
is called the **assignment operator** and it is how you "save" the output of
a function for later use. When the other line of code runs, R performs the
default action of printing the value of `waking_hours` to the screen.

In math, when we have `x`, `y`, etc. as stand-ins for some value, we call `x`
and `y` **variables**. The concept of a variable 
also exists in R, but more often the saved value is referred to as an 
**object**, so I'll favor that terminology. You give an object a short name 
to refer to it easily later. An object can be almost anything, including 
a combination of other objects. The objects we create for now will be simple.

### Your First Object

Fill in the code below with your first name to assign it to `my_name`. 
The first line of code won't print any output—its output just gets assigned to
the R object `my_name`. But once your first name is assigned to `my_name`, 
you can use `my_name` to refer to the value you assigned. Below, we'll use
the `paste()` function to combine a greeting with your name:

```{r my-name, exercise=TRUE}
my_name <- ""
paste("Hi, my name is", my_name)
```

> The `paste()` function simply combines multiple character strings together into
a single character string. The arguments you provide as input can be a mix 
of quoted strings (e.g., `"My name is"`) and variables (also called 
**objects**) you have previously defined.

### Example: Intermediate Objects

To avoid the mess of nesting functions( within functions( within functions)),
we can gradually build up the objects we'll ultimately need to run a desired
function, assigning those objects meaningful names so we can easily use them 
as inputs for our ultimate function call. 

It's easier to understand in an example. Fill in the code below to
programmatically write a sentence that should be motivational but just makes
me want to go back to sleep.

1. Create an object on line 1 — name it `waking_hours` — whose value is 24 minus
the number of hours you normally sleep each night (no one can see this but you, 
so feel free to be silly or reveal your bad habits). 
2. On line 2, create a new object — `waking_minutes` — whose value is
`waking_hours` multiplied by 60 — i.e., the number of minutes you're usually
awake each day. 
3. Line 3 is finished for you. It makes one more object called `sentence` by
combining some text with previously saved values.
4. On line 5, fill in an exclamatory word of your choice inside the quotes.
5. Lastly, on line 6, fill in the arguments to the `paste()` function with
`sentence` and `exclamation` in whatever order feels right to you (separated
by a comma).

```{r fun-sequence, exercise=TRUE}
waking_hours <- 
 <- waking_hours * 60
sentence <- paste("I have", waking_minutes, "minutes to use wisely today.")
exclamation <- paste0("", "!")
paste(, )
```

```{r fun-sequence-hint}
waking_hours <- 6.5
waking_minutes <- waking_hours * 60
sentence <- paste("I have", waking_minutes, "minutes to use wisely today.")
exclamation <- paste0("Golly", "!")
paste(exclamation, sentence)
```

To me, as a human with very human processing limitations, writing out that
sequence is easier to reason through as I'm putting it together than having
to construct the whole thing in one line of code: 
`paste("I have", ((24 - 6.5) * 60), "minutes to use wisely today")`. 
Just as important, when I or someone else reads my code later, it should be 
easier to understand what it means when those intermediate objects are defined
and clearly labeled.

> You are the ruler of your R code universe, so it's your responsibility and 
privilege to name things however you see fit. Instead of naming a dataset
`dat` or `data`, call it what it is: `trial_patients` or `lab_values`, etc.
Try to avoid `patients1`, `patients2`, ... when making a series of changes
to some data — as much as possible choose names that are meaningful 
and intuitive.
## Wrap Up

## Congratulations!

You've finished this tutorial. The concepts we covered here have the 
potential to be boringly basic and/or intimidatingly complex. It's okay if 
you felt either/both of those things. 

We've hammered on the concepts of functions, inputs, outputs, and objects because
from here things will accelerate very quickly. You'll forget about the basics
because you'll be struggling to remember lists of new functions and 
sweating over the technical details of making them work. You won't have time to
remember the theory we've covered, and that's okay. 

One day the stress of memorizing code syntax for specific packages will 
subside and you'll feel curious again about how things work within R as a 
whole system. You can always return to this tutorial, but likely you'll
find more useful resources to learn from. The important thing is that you've
taken the time at least once to build a general mental model of how R works, 
which puts you ahead of many intermediate R users already.

### Recap

Here's a recap of important concepts:

- A **function** takes input (**arguments**) and creates output.
- Everything that happens in R is just functions taking input and 
creating output.
- Most functions live in **packages**, which you can load with the `library()`
function.
- The output from one function can be, and frequently is, the input for another.
- Instead of printing it to the screen or using it in another function right 
away, you can capture the output of a function as an R object, **assigning** it 
a short, meaningful name.
- The **assignment operator** `<-` is how you do that.

### Up Next

Transitioning from here to doing actual analytics work in R is a matter of
temporarily accepting that some things work a certain way and you don't know 
why yet. We'll gloss over details in favor of giving you some 
templates/examples/recipes you can use as jumping off points for your own work.

Important topics to get to next include

- Importing data from files, databases, elsewhere
- Inspecting data for errors or surprises
- Shaping, cleaning, curating, and otherwise wrangling data
- Visualizing data
- Summarizing data
- Modeling data
- Communicating results

R has great tools for all of these things, and they are fun to learn and use.
Stick with it!
